---
title: "Introduction to reconstructr"
author: "Oliver Keyes"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Introduction to reconstructr}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

## Introduction to reconstructr

Within web analytics, session reconstruction is an essential prerequisite of a host of useful and commonly-requested metrics, including time-on-site, time-on-page, and bounce rate. With reconstructr, we can conveniently divide
user data into sessions and calculate a host of metrics from the resulting dataset.

This vignette serves as an introduction to reconstructr, the approach it takes, and the functionality it provides.

### Dividing events into sessions

The nature of a "session" has provided fodder for researchers for years. Most people take an approach based on
inactivity timeouts; if someone does not take an action in N seconds, their session has ended and
a new one begins on their next registered action.

Using reconstructr, we can conveniently divide events into sessions with the `reconstruct_sessions` function. This takes a list of numeric vectors - each one containing the value, in seconds, of each event associated
with a particular user - and divides them into sessions, splitting whenever the time between events is
greater than `threshold` Timestamps can be converted to their representation in seconds with the `to_seconds` function.

```{r, eval=FALSE}
library(reconstructr)
actions_by_user <- list(c(1417330230, 1417330250, 1417330295, 1417330324, 1417330416),
                        c(1417401697, 1417401741, 1417401751, 1417403263))
sessions <- reconstruct_sessions(timestamps = actions_by_user, threshold = 1800)
```

The appropriate value for N - which sessionise represents with the "threshold" argument - has been hotly debated for years, with most systems historically settling on a 30 minute (1800 second) approach. The default for reconstructr is an hour, or 3600 seconds, in line with recent research by  [Halfaker et al.](http://arxiv.org/abs/1411.2878) on the commonality of that threshold to a variety of websites and systems. Really, though, you can set it to whatever value is most appropriate to you: Halfaker et al also demonstrate a good way of working out the appropriate threshold.

### Calculating metrics

Once you have your event dataset divided into sessions, you can easily take advantage of reconstructr to calculate a variety of useful and often-requested metrics, including the length of each session, the bounce rate in the dataset,
and the time between events (or "time on page").

#### Session length

Calculating session length is easy, right? Just work out the time between each pair of events, and add them together.
Except...

1. What do we do about sessions with only one event?
2. How do we work out the time spent idling on the last event, since there's no timestamp to compare it to?

Different people have different approaches to this; Google Analytics, for example, [entirely discounts sessions consisting of single events](https://support.google.com/analytics/answer/1009409?hl=en), and also ignores [time after the last event](https://support.google.com/analytics/answer/1006253?hl=en). Other approaches are to generate
a "padding value" - based on looking at the between-event times that *can* be estimated - and using that for single-event sessions, or for the final event in a session, or both.

Rather than making you pick a methodology you may not agree with, reconstructr allows you to pick your favoured approach and experiment with others, using the ` session_length ` function:

Session length can be calculated using the `session_length` function. With the dataset from above,
as an example:

```{r, eval=FALSE}
length_of_sessions <- session_length(sessions, padding_value = 430, preserve_single_events = FALSE, strip_last = FALSE)
```

With the options present, you can preserve single events (or remove them), strip the last event in a session and ignore it (or not), and modify the padding value to something that best suits your database - which the function `padding_value` can be used to estimate.

#### Events per session

Events per session is a far easier metric to think through, and can be calculated using the `session_events` function:
```{r, eval=FALSE}
events_per_session <- session_events(sessions)
```

This simply gives you a vector, each entry containing a count of how many events were in that session.

#### Bounce rate

The bounce rate is the measure of what proportion of sessions within a dataset consist of only a single event:
in other words, the percentage of sessions in which the user saw one page and, one way or another, left the site.

This can be calculated conveniently with `bounce_rate`:

```{r, eval=FALSE}
bounce_rate <- session_events(sessions, decimal_places = 2)
```

"sessions" refers to the output of a `sessionise` call; "decimal_places" controls how closely the
resulting percentage should be rounded.

#### Time on page

Time on page, or the time between events, is easily calculated by looking at the time between each pair of events
within a session. The reconstructr implementation of this is `event_time` which produces either a vector
or a list of time-on-page values, depending on user choice:

```{r, eval=FALSE}
time_on_page <- event_time(sessions, as_vector = FALSE, fun, ...)
```

Not only can you use this to calculate time-on-page, you can also perform statistical transformations on the resulting data. "fun" takes a function (with `...` used to pass any further arguments to it) which is run over
the data, and the results from that run returned. This means that you can conveniently produce a single number
through calls like:

```{r, eval=FALSE}
average_time_on_page <- event_time(sessions, as_vector = FALSE, mean)
```
